---
title: "Introduction to data"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    css: ../lab.css
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(dplyr)
library(ggplot2)
```

Some define Statistics as the field that focuses on turning information into
knowledge. The first step in that process is to summarize and describe the raw
information - the data. In this lab, you will gain insight into public health
by generating simple graphical and numerical summaries of a data set collected 
by the Centers for Disease Control and Prevention (CDC). As this is a large 
data set, along the way you'll also learn the indispensable skills of data 
processing and subsetting.

## Getting started

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone 
survey of 350,000 people in the United States. As its name implies, the BRFSS 
is designed to identify risk factors in the adult population and report 
emerging health trends. For example, respondents are asked about their diet and 
weekly physical activity, their HIV/AIDS status, possible tobacco use, and even
their level of healthcare coverage. The BRFSS Web site 
([http://www.cdc.gov/brfss](http://www.cdc.gov/brfss)) contains a complete 
description of the survey, including the research questions that motivate the 
study and many interesting results derived from the data.

We will focus on a random sample of 20,000 people from the BRFSS survey 
conducted in 2000. While there are over 200  variables in this data set, we will
work with a small subset.

We begin by loading the data set of 20,000 observations into the R workspace. 
After launching RStudio, enter the following command.

```{r load-data}
load(url("https://stat.duke.edu/~mc301/data/cdc.RData"))
```

The data set `cdc` that shows up in your workspace is a *data matrix*, with each
row representing a *case* and each column representing a *variable*. R calls 
this data format a *data frame*, which is a term that will be used throughout 
the labs.

To view the names of the variables, type the command

```{r names}
names(cdc)
```

This returns the names `genhlth`, `exerany`, `hlthplan`, `smoke100`, `height`, 
`weight`, `wtdesire`, `age`, and `gender`. Each one of these variables 
corresponds to a question that was asked in the survey. For example, for 
`genhlth`, respondents were asked to evaluate their general health, responding
either excellent, very good, good, fair or poor. The `exerany` variable 
indicates whether the respondent exercised in the past month (1) or did not (0).
Likewise, `hlthplan` indicates whether the respondent had some form of health 
coverage (1) or did not (0). The `smoke100` variable indicates whether the 
respondent had smoked at least 100 cigarettes in her lifetime. The other 
variables record the respondent's `height` in inches, `weight` in pounds as well
as their desired weight, `wtdesire`, `age` in years, and `gender`.

A very useful function for taking a quick peek at your dataset, and viewing
its dimensions and data types is `str`.

```{r cdc}
str(cdc)
```

Note that R calls categorical variables `factor`s.

1.  How many cases are there in this data set? How many variables? For each 
    variable, identify its data type (e.g. categorical: ordinal or not,
    numerical: continuous or discrete). Do not just rely on the R output,
    also think about the nature of the variables.

## Describing distributions

### Histograms

The BRFSS questionnaire is a massive trove of information. A good first step in
any analysis is to distill all of that information into a few summary statistics
and graphics. 

Let's start with some histograms. We can create a the histogram for the age of
our respondents with the following command:

```{r hist-age-geom}
qplot(x = age, data = cdc, geom = "histogram")
```

This functions says to plot `age` on the x-axis. It also defines a `geom` (short for 
geometric object), which describes the type of plot you will produce. 

Actually, since the default `geom` for a single numerical variable is a histogram, 
we could get away with not defining a `geom` and still get the same plot:

```{r hist-age}
qplot(x = age, data = cdc)
```

It's a good idea to be explicit in our code and always define all options (and not
rely on default settings), especially when we're first learning a new language.

Histograms are generally a very good way to see the shape of a single 
distribution, but that shape can change depending on how the data is split 
between the different bins. You can easily define the binwidth you want to use:

```{r hist-bmi}
qplot(x = age, data = cdc, geom = "histogram", binwidth = 1)
qplot(x = age, data = cdc, geom = "histogram", binwidth = 20)
```

How do these histograms with the various binwidths compare?

### Summary statistics

As a simple example, the function `summary` returns a numerical 
summary: minimum, first quartile, median, mean, third quartile, maximum.

To calculate the summary statistics for `weight`, type

```{r summary-weight}
cdc %>%
  select(weight) %>%
  summary()
```

<div id="boxedtext">
**A note on piping:** Note that we can read these three lines of code as the following: 

*"Take the `cdc` dataset and **pipe** it into the `select` function. Using this 
function select the variable called `weight`, and **pipe** this 
variable into the `summary` function."*

The `%>%` operator is called the **piping** operator. Basically, it takes the output
of the current line and pipes it into the following line of code.
</div>

Since R also functions like a very fancy calculator, you can use these statistics
to then calculate the interquartile range for the respondents' weight, as the value
of Q3 - Q1.

```{r weight-range-arith}
190 - 140
```

You can also individually calculate summary statistics, and make your own customized list
of them using the `summarise` function:

```{r summary-custom-list}
cdc %>%
  summarise(mean_wt = mean(weight), sd_wt = sd(weight), mean_ht = mean(height), sd_ht = sd(height))
```

Note that in the `summarise` function we created a list of four elements. The names
of these elements are user defined, like `mean_wt`, `sd_wt`, etc. and you could
customize these names as you like (just don't use spaces in your names). Calculating
these summary statistics also require that you know the function calls. Some useful
function calls for summary statistics for a single numerical variable are as follows:

- `mean`
- `median`
- `sd`
- `var`
- `IQR`
- `range`
- `min`
- `max`

Another useful functionality is being able to quickly calculate summary statistics for
various groups in your dataset. For example, we can modify the above command using the
`group_by` function  to get the same summary stats for males and females:

```{r summary-custom-list-gender}
cdc %>%
  group_by(gender) %>%
  summarise(mean_wt = mean(weight), sd_wt = sd(weight), mean_ht = mean(height), sd_ht = sd(height))
```

Here, we first grouped the data by gender, and then calculated the summary statistics.

2.  Calculate the median and interquartile range for `weight`s of people who
    did and did not exercise in the last month.
    
### Tables

While it makes sense to describe a quantitative variable like `weight` in terms
of these statistics, what about categorical data? We would instead consider the
sample frequency or relative frequency distribution. The function `table` does
this for you by counting the number of times each kind of response was given.
For example, to see the number of people who have smoked 100 cigarettes in their
lifetime, type

```{r table-smoke}
cdc %>%
  select(smoke100) %>%
  table()
```

or instead look at the relative frequency distribution by typing

```{r table-smoke-prop}
cdc %>%
  select(smoke100) %>%
  table()/20000
```

Notice how R automatically divides all entries in the table by 20,000 in the 
command above. 

To make a bat plot of these data, use the following:

```{r table-smoke-barplot}
qplot(x = smoke100, data = cdc, geom = "bar")
```

3.  Compute the relative frequency distribution for `gender` and also for `genhlth`. 
    How many males are in the sample? What proportion of the sample reports 
    being in excellent health?

The `table` command can be used to create contingency tables as well. 
For example, to examine which participants have smoked across each 
gender, we could use the following.

```{r table-smoke-gender}
cdc %>%
  select(gender, smoke100) %>%
  table()
```

4. **a.** What percent of males are smokers? What percent of females are smokers? 
**b.** What percent of the sample are males who are smokers? What percent of the sample 
are females who are smokers? **c.** Which pair of statistics is more useful for 
determining whether males or females are more likely to be smokers? Explain your 
reasoning.
    
## Visualizing relationships

### Between two numercial variables

In the last lab we visualized relationships between two numerical variables using
scatterplots. As a quick reminder, let's make one more of those:

```{r height-weight}
qplot(x = weight, y = height, data = cdc, geom = "point")
```

How are these two variables related?

### Between two categorical variables

To create a segmented bar plots we can use the following two optoons.

```{r mosaic-smoke-gender}
qplot(x = gender, fill = smoke100, data = cdc, geom = "bar")
qplot(x = gender, fill = smoke100, data = cdc, geom = "bar", position = "fill")
```

5.  How are the two plots different from each other? Which one is more
    useful for comparing the *proportions* of male and female smokers?

### Between a numerical and a categorical variable

The purpose of a boxplot is to provide a thumbnail sketch of a variable for the purpose 
of comparing across several categories. So we can, for example, compare the heights of 
men and women with 

```{r boxplot-height-gender}
qplot(y = height, x = gender, data = cdc, geom = "boxplot")
```

The notation here is new. `geom`, short for geometric objects, describe the type of 
plot you will produce. So we're asking R to give us a box plots of heights where 
the groups are defined by gender. 

Next let's consider a new variable that doesn't show up directly in this data 
set: Body Mass Index (BMI) 
([http://en.wikipedia.org/wiki/Body_mass_index](http://en.wikipedia.org/wiki/Body_mass_index)). 
BMI is a weight to height ratio, and can be calculated as:

\[ BMI = \frac{weight~(lb)}{height~(in)^2} * 703 \]

703 is the approximate conversion factor to change units from metric (meters and 
kilograms) to imperial (inches and pounds).

We can use the `mutate` function to create this new variable and add it to the
`cdc` dataset:

```{r bmi}
cdc <- cdc %>%
  mutate(bmi = (weight / height^2) * 703)
```

Notice that `(weight / height^2) * 703` is just some arithmetic, but it's applied to 
all 20,000 values in the dataset. That is, for each of the 20,000 
participants, we take their weight, divide by their height-squared and then 
multiply by 703. The result is 20,000 BMI values, one for each respondent. This 
is one reason why we like R: it lets us perform computations like this using 
very simple expressions.

Now we can use this new variable in our analysis. For example, let's create side-by-side 
box plots of `bmi` over the levels of `genhlth`:

```{r boxplot-bmi-genhlth}
qplot(y = bmi, x = genhlth, data = cdc, geom = "boxplot")
```

Discuss what this box plot shows in your team.

6.  Pick another categorical variable from the data set and see how it relates to BMI.
    List the variable you chose, why you might think it would have a relationship to 
    BMI, create side-by-side box plots of the distribution of BMI over this variable,
    and provide an interpretation for this plot.

## Subsetting

It's often useful to extract all individuals (cases) in a data set that have 
specific characteristics. We can do this easily using the `filter` function and
a series of **logical operators**. The most commonly used logical operators for
data analysis are

- `==` means "equal to"
- `!=` means "not equal to"
- `>` or `<` means "greater than" or "less than"
- `>=` or `<=` means "greater than or equal to" or "less than or equal to"

Using these, we can create a subset of the `cdc` dataset for just the men, and save 
this as a new dataset called `males`:

```{r males}
males <- cdc %>%
  filter(gender == "m")
```

Here, we've created a new object, called `males`. The special symbol `<-` 
performs an *assignment*, taking the output of one line of code and saving it 
into this new object that you defined.

We can take look at the contents of this object by clicking on it in the 
*Environment* tab, or by typing the following command we can view the first few 
lines of it

```{r head-males}
head(males)
```

We don't want to print the entire dataset since it's pretty large, taking
a look at just the first few lines should give you a good idea of what it
looks like.

This new data set contains all the same variables but just under half the rows.
It is also possible to tell R to keep only specific variables using the `select`
function we learned earlier, but this is not relevant here since we are focusing
on subsetting a dataset based on values of one or more variables.

As an aside, you can use several of these conditions together with `&` and `|`.
The `&` is read "and" so that

```{r males-and-over30}
males_and_over30 <- cdc %>%
  filter(gender == "m" & age > 30)
```

will give you the data for men over the age of 30. The `|` character is read 
"or" so that

```{r males-or-over30}
males_or_over30 <- cdc %>%
  filter(gender == "m" | age > 30)
```

will take people who are men or over the age of 30 (why that's an interesting 
group is hard to say, but right now the mechanics of this are the important 
thing). In principle, you may use as many "and" and "or" clauses as you like 
when forming a subset.

7.  Create a new object called `under23_and_smoke` that contains all observations 
    of respondents under the age of 23 that have smoked 100 cigarettes in their 
    lifetime. Write the command you used to create the new object as the answer
    to this exercise, and report the number of cases that meet this criteria.
    
## Recap

At this point, we've done a good first pass at analyzing the information in the 
BRFSS questionnaire. We've found an interesting association between smoking and
gender, and we can say something about the relationship between people's 
assessment of their general health and their own BMI. We've also picked up 
essential computing tools -- summary statistics, subsetting, and plots -- that 
will serve us well throughout this course.

* * *

<div class="oyo">

## On Your Own

-   Make a scatterplot of weight versus desired weight. Describe the 
    relationship between these two variables.

-   Let's consider a new variable: the difference between desired weight 
    (`wtdesire`) and current weight (`weight`). Create this new variable by 
    subtracting the two columns in the data frame and assigning them to a new 
    variable in the `cdc` dataset called `wdiff`.

-   What type of data is `wdiff`? If an observation `wdiff` is 0, what does 
    this mean about the person's weight and desired weight. What if `wdiff` is 
    positive or negative? *Hint:* This is the only exercise in this lab that
    doesn't require code to answer.

-   Describe the distribution of `wdiff` in terms of its center, shape, and 
    spread, including any plots and numerical summaries you use. What does this 
    tell us about how people feel about their current weight?

-   Using numerical summaries and a side-by-side box plots, determine if men tend
    to view their weight differently than women.

-   Now it's time to get creative. Find the mean and standard deviation of 
    `weight` and determine what proportion of the weights are within one 
    standard deviation of the mean.

</div>

<div id="license">
This is a product of OpenIntro that is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel 
from a lab written by Mark Hansen of UCLA Statistics.
</div>