---
title: 'Foundations for statistical inference - Confidence intervals'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

## Sampling from Ames, Iowa

If you have access to data on an entire population, say the size of every 
house in Ames, Iowa, it's straight forward to answer questions like, "How big 
is the typical house in Ames?" and "How much variation is there in sizes of 
houses?". If you have access to only a sample of the population, as is often 
the case, the task becomes more complicated. What is your best guess for the 
typical size if you only know the sizes of several dozen houses? This sort of 
situation requires that you use your sample to make inference on what your 
population looks like.

## The data

In the previous lab, "Sampling Distributions", we looked at the population data
of houses from Ames, Iowa. Let's start by loading that data set.

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(eval = FALSE)
```

```{r load-data, message=FALSE}
require(mosaic)
require(oilabs)
data(ames)
head(ames)
```

We can rename the `Gr.Liv.Area` variable in the same way we did in the last lab, for consistency.

```{r rename}
ames <- ames %>%
  rename(area = Gr.Liv.Area)
```

In this lab we'll start with a simple random sample of size 60 from the 
population. Note that 
the data set has information on many housing variables, but for the first 
portion of the lab we'll focus on the size of the house, represented by the 
variable `area`.

```{r sample}
population <- ames %>% 
  select(area) 
samp <- sample(population, 60)
```

1.  Describe the distribution of your sample. What would you say is the 
"typical" size within your sample? Also state precisely what you interpreted 
"typical" to mean.

2.  Would you expect another student's distribution to be identical to yours? 
Would you expect it to be similar? Why or why not?

## Confidence intervals

One of the most common ways to describe the typical or central value of a 
distribution is to use the mean. In this case we can calculate the mean of the 
sample using,

```{r sample-mean}
sample_mean <- mean(~area, data = samp)
```

Return for a moment to the question that first motivated this lab: based on 
this sample, what can we infer about the population? Based only on this single 
sample, the best estimate of the average living area of houses sold in Ames 
would be the sample mean, usually denoted as $\bar{x}$ (here we're calling it 
`sample_mean`). That serves as a good *point estimate* but it would be useful 
to also communicate how uncertain we are of that estimate. This can be 
captured by using a *confidence interval*.

We can calculate a 95% confidence interval for a sample mean by adding and 
subtracting a particular number of standard errors. You might have that number memorized 
for the 95% confidence interval, but lets use `qnorm()` to confirm.

```{r zstar}
zstar <- qnorm(0.975) # why 0.975?
zstar
```

Now we can build the confidence interval. 

```{r ci}
se <- sd(~area, data = samp) / sqrt(60)

lower <- sample_mean - zstar * se

upper <- sample_mean + zstar * se

c(lower, upper)
```

This is an important inference that we've just made: even though we don't know 
what the full population looks like, we're 95% confident that the true 
average size of houses in Ames lies between the values *lower* and *upper*. 
There are a few conditions that must be met for this interval to be valid.

3.  For the confidence interval to be valid, the sample mean must be normally 
distributed and have standard error $s / \sqrt{n}$. What conditions must be 
met for this to be true?

## Confidence levels

4.  What does "95% confidence" mean? If you're not sure, see Section 2.8.4.

In this case we have the luxury of knowing the true population mean since we 
have data on the entire population. This value can be calculated using the 
following command:

```{r pop-mean}
mean(~area, data = population)
```

5.  Does your confidence interval capture the true average size of houses in 
Ames? If you are working on this lab in a classroom, does your neighbor's 
interval capture this value? 

6.  Each student in your class should have gotten a slightly different 
confidence interval. What proportion of those intervals would you expect to 
capture the true population mean? Why? If you are working in this lab in a 
classroom, collect data on the intervals created by other students in the 
class and calculate the proportion of intervals that capture the true 
population mean.

Using R, we're going to recreate many samples to learn more about how sample 
means and confidence intervals vary from one sample to another. The *do()* function comes 
in handy here (If you are unfamiliar with `do()`, review the [Sampling Distribution Lab](http://htmlpreview.github.io/?https://github.com/beanumber/oiLabs-mosaic/blob/master/sampling_distributions/sampling_distributions.html)).

Here is the rough outline:

-   Obtain a random sample.
-   Calculate and store the sample's mean and standard deviation.
-   Repeat steps (1) and (2) 50 times.
-   Use these stored statistics to calculate many confidence intervals.

We can accomplish this using the `do` and `favstats` functions in `mosaic`. The following lines of code take a uniform random `sample` of size 60 from `population`, computes several summary statistics (including `mean` and `sd`), and then does this 50 times and saves the result as a data frame. 

```{r loop, tidy = FALSE}
n <- 60
samp <- do(50) * favstats(~area, data = sample(population, size = n))
```

Lastly, we construct the confidence intervals.

```{r ci50}
samp <- samp %>%
  mutate( se = sd / sqrt(n)) %>%
  mutate( lower = mean - zstar * se) %>%
  mutate( upper = mean + zstar * se)
```

Lower bounds of these 50 confidence intervals are stored in `lower`, and the 
upper bounds are in `upper`. Let's view the first interval.

```{r first-interval}
samp %>% 
  select(upper, lower) %>% 
  slice(1)
```

Now, we can plot how many of our confidence intervals intersect the true population mean using the `plot_ci` function.

```{r plot-ci}
with(samp, plot_ci(lower, upper, m = mean(~area, data = population)))
```

7. How many of your confidence intervals contained the true mean? Is this the number you expected to see?

* * *

## On your own

So far, we have only focused on estimating the mean living area in homes in Ames.  Now you'll try to estimate the mean home price. 

- Create 50 random samples of size 60. Generate 95% confidence intervals for each of your samples, and plot the confidence intervals with the `plot_ci` function.  What proportion of your confidence intervals include the true population mean of `home price`? Is this proportion exactly equal to the confidence level? If not, explain why.

```{r plot-ci2, eval=FALSE}
with(samp, plot_ci(lower, upper, m = mean(~area, data = population)))
```

- Pick a confidence level of your choosing, provided it is not 95% or 99%. What is the appropriate critical value?

- Calculate 50 confidence intervals at the confidence level you chose in the previous question. You do not need to obtain new samples, simply calculate new intervals based on the sample means and standard deviations you have already collected. Using the *plot_ci* function, plot all intervals and calculate the proportion of intervals that include the true population mean of `home price`. How does this percentage compare to the confidence level level selected for the intervals?


<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was written for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel.
</div>
